Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-03-31 doc/thesis/conclusion.tex
Finished results! Yes!

>>>>

Our use of the \gls{BA} model makes several simplifying assumptions. First, we
assume homogeneity across the network with respect to node behaviour and
transmission risk. In reality, the attraction to high-degree nodes seems likely
to vary among individuals, as does their risk of transmitting or contracting
the virus. We have also assumed that all transmission risks are symmetric,
which is clearly false for all known modes of \gls{HIV} transmission, and that
infected individuals never recover but remain infectious indefinitely. These
assumptions were made for the purpose of keeping the model as simple as
possible, since this is the very first attempt to fit a contact network model
in a phylodynamic context. However, the Gillespie simulation algorithm built
into \software{netabc} can handle arbitrary transmission and removal rates
which need not be homogeneous across the network. Moreover, it is possible to
use kernel-\gls{ABC} to fit a model which relaxes some or all of these
assumptions, which may be a fruitful avenue for future investigation.

### 2016-03-31 doc/thesis/methods.tex
Finished results! Yes!

\subsection{Computer program}

\software{Netabc} is a computer program designed to perform statistical
inference of contact network parameters from an estimated transmission tree
using kernel-\gls{ABC}. The program combines three major components: Gillespie
simulation, to simulate transmission trees on contact networks; the
phylogenetic kernel, to compare simulated to observed transmission trees; and
adaptive \gls{ABC}-\gls{SMC}, to maintain a population of particles and
advance it toward the \gls{ABC} target distribution. We give a high-level
overview of the program here, before describing these three components in
detail.

@@

\software{Netabc} is written in the \software{C} programming language. The
\software{igraph} library~\autocite{csardi2006igraph} is used to generate and
store contact networks and phylogenies. Judy arrays~\autocite{baskins2004judy}
are used for hash tables and dynamic programming matrices. The
\gls{GSL}~\autocite{gough2009gnu} is used to generate random draws from
probability distributions, and to perform the bisection step in the adaptive
\gls{ABC}-\gls{SMC} algorithm. Parallelization is implemented with POSIX
threads~\autocite{barney2009posix}. In addition to the \software{netabc} binary
to perform kernel-\gls{ABC}, we provide three additional stand-alone utilities:
\software{treekernel}, to calculate the phylogenetic kernel;
\software{nettree}, to simulate a transmission tree over a contact network; and
\software{treestat}, to compute various summary statistics of phylogenies. The
programs are freely available at \url{github.com/rmcclosk/netabc}.

@@

The algorithm may be stopped when the tolerance reaches a
user-defined final value, or when the rate of acceptance of the
Metropolis-Hastings kernel reaches a user-defined threshold. Following a
heuristic applied by the authors~\autocite{del2012adaptive}, we used the latter
stopping criterion, accepting the SMC approximation to the posterior when the
\gls{MCMC} acceptance rate dropped below 1.5\%.

\subsection{Kernel classifiers}
\label{subsec:kernel}

Recall that our \gls{ABC} approach to fitting contact network models involves
simulating transmission trees under a wide variety of parameter values, and
then comparing these simulated trees to the true transmission tree. Values
which produce trees similar to the observed transmission tree are distinguished
as more likely than values which produce trees very different from the truth.
In order for this type of analysis to succeed, it is critical that different
parameter values produce different looking trees. Otherwise, if many different
values produce trees which are too similar to each other, it will be impossible
to distinguish which value is most consistent with the real tree. Just as
importantly, trees simulated with similar parameter values must be similar to
each other. In mathematical terms, we require the trees simulated from distinct
parameter values to be \defn{separable} in tree space. The concept of
separability is illustrated in Figure~\ref{fig:separable}.

\begin{figure}[ht]
  \centering
  \includegraphics{separable.pdf}
  \caption[Separable versus non-separable parameters in kernel space]{
    Separable versus non-separable parameters in kernel space. Trees have been
    simulated under three sets of parameters, represented as blue, red, and
    green. An observed tree is shown in black. Trees are laid out such that
    the distance between two trees corresponds to their similarity. In panel A,
    trees from the same parameter set are similar to each other, but different
    from other trees. The true tree is most consistent with the red parameters.
    In panels B and C, trees from the same parameter set are not similar to
    each other (B), or trees from different parameter sets are similar to each
    other (C). It's difficult to say which parameters the true tree is most
    consistent with.
  }
  \label{fig:separable}
\end{figure}

Before undertaking a complete ABC analysis, I analyzed the parameters of the
\gls{BA} network model to to determine whether they could be separated in tree
kernel space. A graphical schematic of the analysis undertaken here is given in
Figure~\ref{fig:kernelexpt}.

\begin{figure}[ht]
  \centering
  \label{fig:kernelexpt}
  \includegraphics{kernel-expt.pdf}
  \caption[Schematic of first set of simulation experiments]{
    Schematic of first set of simulation experiments to determine separable
    parameters in kernel space, along with optimal tree kernel meta-parameters.
  }
\end{figure}

The method of testing for separability was described previously
in~\autocite{poon2015phylodynamic}. As a concrete example, consider the
attachment power parameter \gls{alpha}. This parameter describes the strength
of attraction to highly connected nodes and is bounded below by zero indicating
no extra attraction. By qualitative observation, I determined that a power of
2.0, which produced networks with very few ``hub'' nodes with extremely high
degree, was a suitable upper bound (\cref{fig:alphabds}). Therefore,
I chose to test the values 0.5, 1.0, and 1.5 for separability. The other
parameters were fixed: the number of nodes in the network was 5000, and the
mean degree of each node in the network was four. As discussed in
subsection~\ref{subsubsec:generative}, this is the smallest mean degree value
for preferential attachment networks which produces networks which are more
than trees.

\begin{figure}[ht]
  \centering
  \includegraphics{alpha-bounds.pdf}
  \caption[Upper and lower bounds on preferential attachment power]{
    Qualitative justification for choice of zero and two as lower and upper
    bounds on preferential attachment power $\alpha$. (A) Preferential
    attachment network on 50 nodes with $\alpha = 0$, the lower bound enforced
    by the model. (B) Preferential attachment network on 50 nodes with $\alpha
    = 2$, where a few nodes have very high degree but the majority have very
    low degree. (C) Density plot of node degrees in a 5000-node network with
    $\alpha = 0$. The maximum degree of a node in this network was 24. (D)
    Density plot of node degrees in a 5000-node network with $\alpha = 2$. The
    maximum degree of a node in this network was 4942. 
  }
  \label{fig:alphabds}
\end{figure}

For each of the values of $\alpha$, I generated 100 networks, each with 5000
nodes. An epidemic was simulated over each network (see
\cref{subsubsec:nettree}) until 1000 nodes were infected, and 500 of those
infected nodes were sampled to form a transmission tree. This resulted in 300
total simulated transmission trees - 100 for each of the three values of
$\alpha$. The data generation steps are shown on the left side of
Figure~\ref{fig:kernelexpt}. Next, I computed the tree
kernel~\autocite{poon2013mapping} (see subsection~\ref{subsubsec:treeshape})
for each pair of trees. These values were placed into a 300 $\times$ 300 kernel
matrix, where the value at the $(i, j)$th position was the tree kernel of the
$i$th and $j$th trees.

The tree kernel provides a pairwise similarity score between two trees. The
higher the kernel score, the more similar the trees are to each other.
Therefore, to have separability, we need trees simulated with the same value of
$\alpha$ to have high kernel scores with each other, but low kernel scores with
trees from different $\alpha$ values. We can visually check whether or not this
is true by laying out the trees as points on a graph, in such a way that trees
with high scores are close to each other, but trees with low scores are far
apart. This is accomplished by performing a kernel principal components
analysis (kPCA)~\autocite{scholkopf1998nonlinear} on the kernel matrix.
Briefly, ordinary principal components analysis (PCA) finds a lower dimensional
representation of points in a high dimensional space which preserves as much of
the variation in the data as possible. kPCA performs the same task, but using
the dot products of each pair of points as input, instead of the data points
themselves. A two-dimensional kPCA projection of the simulated trees is shown
in Figure~\ref{fig:pakpca}. The scenario just described - 500 samples from 1000
infected nodes - is the central panel. To ensure that the method could be used
in a variety of contexts, the same analysis was performed with 500 and 2000
infected nodes, as well as 100 and 1000 sampled tips.

The next step was to quantify how well trees simulated with different $\alpha$
values could be distinguished from each other. As described
previously~\autocite{poon2015phylodynamic}, this was done by assessing the
accuracy of a support vector machine regression
(SVR)~\autocite{smola1997support}. Briefly, an SVR operates by finding a
hyperplane (in two dimensions, a line) such that the deviation of most of the
data points from the line is less than some prescribed threshold. Points
further away than this threshold are ignored in the model. I performed 1000
replicate 2-fold cross-validations of an SVR predicting preferential attachment
power on the simulated trees, using the \software{ksvm} function from
\software{kernlab}~\autocite{karatzoglou2004kernlab}. That is, the SVR was
trained on a random subset of 150 trees, and then used to predict $\alpha$ of
the remaining 150 trees. The predictions were correlated against the true
values of $\alpha$ to obtain an $R^2$, and this procedure was repeated 1000
times with different subsets of trees.

The cross-validation had the dual purpose of providing a means to selecting the
optimal meta-parameters to the tree kernel (see
subsection~\ref{subsubsec:treeshape}) - they are be those which provide the
highest average $R^2$. To this end, the cross-validation was repeated for
several values of $\lambda$ and $\sigma$ as shown in Figure~\ref{fig:kernelexpt}
Each combination was evaluated with and without multiplying the tree kernel by
the normalized lineages-through-time (nLTT) statistic, and was further repeated
for the scenarios with differing numbers of infected and sampled nodes
described above. To ensure that the tree kernel was the most appropriate
similarity measure to use for ABC, we also computed the $R^2$ of $\alpha$
against Sackin's index, a widely used tree balance statistic (see
subsection~\ref{subsubsec:treeshape}), by the same cross-validation procedure.

\subsection{Grid search}

The previous set of simulations were intended to investigate which contact
network parameters could and could not be inferred by examining transmission
trees. However, they tell us nothing about the accuracy or precision we might
expect when inferring those parameters numerically. As illustrated in
Figure~\ref{fig:accprec}, the ideal situation is one where we are accurate and
precise, and the worst situation is when we are precise but not accurate.

\begin{figure}[ht]
  \centering
  \includegraphics{acc-prec.pdf}
  \caption{Illustration of accurate vs. precise kernel score estimates}
  \label{fig:accprec}
\end{figure}

Figure~\ref{fig:gridsearch} shows a schematic of this experiment. A number of
representative values, here denoted $k$, were chosen for the parameter of
interest. In the case of preferential attachment power $\alpha$, I chose $k =
8$ testing values, namely $0, 0.25, \ldots, 2.0$. For each of these values, ten
networks on 5000 nodes were generated, an epidemic of 1000 nodes was simulated
over each, and a transmission tree of 500 tips was sampled. The resulting $10
\times k$ simulated transmission trees were referred to as the \defn{testing
trees}. In addition, I chose a further $n \gg k$ training values spanning the
range of the parameter. For $\alpha$, these were $0, 0.01, \ldots, 2.0$.
Fifteen trees were simulated in the same manner for each of these values,
referred to here as \defn{training trees}.

For each of the testing trees, I computed the tree kernel with all of the
training trees. This resulted in 15 kernel scores per training value. The
training value with the highest median kernel score was used as a point
estimate of the testing value. The kernel scores were then normalized to lie in
[0, 1], and an interval was found which contained 95\% of the area under the
curve and was of minimal width. This was used as a 95\% confidence interval for
the parameter.

\begin{figure}[ht]
  \centering
  \includegraphics{gridsearch-expt.pdf}
  \caption{Schematic of grid search simulation experiments.}
  \label{fig:gridsearch}
\end{figure}

\subsection{Approximate Bayesian computation}

Our final set of simulation experiments was to test the ABC-SMC algorithm on
simulated data. For each replicate, we generated a network on 5000 nodes,
simulated an epidemic over that network until 1000 nodes were infected, and
sampled either 500 or all 1000 of those nodes to form a transmission tree.
Priors were specified as follows: for the total number of nodes in the network,
$\Uniform(1000, 10000)$; for the number of infected nodes, $\Uniform(500,
2000)$; for the preferential attachment power, $\Uniform(0, 2)$. The mean
degree of nodes in the network was fixed at either 4, 10, or 16. The
algorithm was run with 1000 particles, and was stopped when the \gls{MCMC}
acceptance probability dropped below 1.5\%.

\subsection{Characterization of power-law exponent in Barab\'asi-Albert networks}

Most studies of social network or transmission network
parameters~\autocite[e.g.][]{liljeros2001web, jones2003assessment,
schneeberger2004scale, brown2011transmission} report the coefficient
\gls{gamma} of the power law degree distribution. To make our results
comparable to previous work, we used simulated networks to investigate the
relationship between the \gls{BA} model parameters and \gls{gamma}. A network
was simulated for each element of the Cartesian product of these parameters:
\gls{N} = \sett{500, 600, \ldots, 15000}, \gls{m} = \sett{1, 2, \ldots, 8}, and
\gls{alpha} = \sett{0.00, 0.01, \ldots, 2.00}. A power law distribution was
fitted to the degree distribution of each simulated network using the
\software{fit\_power\_law} function in \software{igraph} with the `R.mle'
implementation. We fitted a \gls{GLM} with Gamma-distributed errors and a log
link function to the observed distribution of \gls{gamma} values, with
\gls{alpha}, \gls{m}, \gls{N}, and all possible interaction terms as
predictors.

@@

We applied kernel-\gls{ABC} to six \gls{HIV} datasets, summarized in
\cref{tab:realdata}.

\begin{table}
  \centering
  \input{\tablepath/realdata-bc}
  \caption{Characteristics of published HIV datasets analyzed with kernel-ABC.}
  \label{tab:realdata}
\end{table}

>>>>

\subsection{Kernel-ABC method}

I wrote a computer program called \software{netabc} to perform statistical
inference of contact network parameters from an estimated transmission tree
using kernel-\gls{ABC}. The program combines three major components: Gillespie
simulation, to simulate transmission trees on contact networks; the
phylogenetic kernel, to compare simulated to observed transmission trees; and
adaptive \gls{ABC}-\gls{SMC}, to maintain a population of particles and advance
it toward the \gls{ABC} target distribution. We give a high-level overview of
the program here, before describing these three components in detail.

@@

\software{Netabc} is written in the \software{C} programming language. The
\software{igraph} library~\autocite{csardi2006igraph} is used to generate and
store contact networks and phylogenies. Judy arrays~\autocite{baskins2004judy}
are used for hash tables and dynamic programming matrices. The
\gls{GSL}~\autocite{gough2009gnu} is used to generate random draws from
probability distributions, and to perform the bisection step in the adaptive
\gls{ABC}-\gls{SMC} algorithm. Parallelization is implemented with POSIX
threads~\autocite{barney2009posix}. In addition to the \software{netabc} binary
to perform kernel-\gls{ABC}, we provide three additional stand-alone utilities:
\software{treekernel}, to calculate the phylogenetic kernel;
\software{nettree}, to simulate a transmission tree over a contact network; and
\software{treestat}, to compute various summary statistics of phylogenies. The
programs are freely available at \url{https://github.com/rmcclosk/netabc}.

@@

The algorithm may be stopped when the tolerance reaches a user-defined final
value, or when the rate of acceptance of the \gls{MH} kernel reaches a
user-defined threshold.

\subsection{Analysis of \acrlong{BA} model}

We investigated four parameters related to the \gls{BA} model, denoted
\gls{N}, \gls{m}, \gls{alpha}, \gls{I}. The first three of these are parameters
of the model itself, while \gls{I} is related to the simulation of transmission
trees over the network. However, we will refer to all four as \gls{BA}
parameters. \gls{N} denotes the total number of nodes in the network, or
equivalently, susceptible individuals in the population. When a node is added
to the network, \gls{m} new undirected edges are added incident to it, and are
attached to existing nodes of degree $k$ with probability proportional to
$k^\alpha$ (\cref{subsec:pa}). To simulate transmission trees over a
\gls{BA} network, we allowed an epidemic to spread until \gls{I} nodes were
infected, and sampled a transmission tree at that time. We assumed that all
contacts had symmetric transmission risk, which was implemented by replacing
each undirected edge in the network with two directed edges (one in each
direction).

We did not consider the time scale of the transmission trees in these
simulations, only their shape. Therefore, the transmission rate along each edge
in the network was set to 1, and all transmission trees' branch lengths were
scaled by their mean. The removal rate of each node was set to 0, implying no
recovery or death in the population. These assumptions are similar to those
made by \textcite{leventhal2012inferring}.

\subsubsection*{Kernel classifiers}
\label{subsec:kernel}

The experiments presented here involved a large number of variables which were
varied combinatorially. For ease of exposition, we will describe a single
experiment first, then enumerate the values of all variables for which the
experiment was repeated. The parameters of the tree kernel, $\lambda$ and
$\sigma$ (\cref{subsubsec:treeshape}) will be referred to as
\defn{meta-parameters} to distinguish them from the parameters of the \gls{BA}
model. With the exception of our own binaries, all analyses were done in
\software{R}, and all packages listed below are \software{R} packages.

The attachment power parameter \gls{alpha} was varied among three values: 0.5,
1.0, and 1.5. For each value, the \software{sample\_pa} function in the
\software{igraph} package was used to simulate 100 networks, with the other
parameters set to \gls{N} = 5000 and \gls{m} = 2. This step yielded a total of
300 networks. An epidemic was simulated on each network using our
\software{nettree} binary until \gls{I} = 1000 nodes were infected, at which
point 500 of them were sampled to form a transmission tree. A total of 300
transmission trees were thus obtained, comprised of 100 trees for each of the
three values of \gls{alpha}. The trees were ``ladderized'' so that the subtree
descending from the left child of each node was not smaller than that
descending from the right child. Summary statistics, such as Sackin's index and
the ratio of internal to terminal branch lengths, were computed for each
simulated tree using our \software{treestat} binary. The trees were visualized
using the \software{ape} package. Our \software{treekernel} binary was used to
calculate the value of the kernel for each pair of trees, with the
meta-parameters set to $\lambda = 0.3$ and $\sigma = 4$. These values were
stored in a symmetric 300 $\times$ 300 kernel matrix. Similarly, we computed
the \gls{nltt} statistic between each pair of trees using our
\software{treestat} binary, and stored them in a second $300 \times 300$
matrix.

To investigate the effect of \gls{alpha} on tree shape, we constructed
classifiers for \gls{alpha} based on three statistics. First, we used the
\software{kernlab} package~\autocite{karatzoglou2004kernlab} to create a
\gls{kSVM} classifier using the computed kernel matrix. Second, we used the
\software{e1071} package to create an ordinary \gls{SVM} classifier using the
pairwise \gls{nltt} matrix. Finally, we performed an ordinary linear regression
of \gls{alpha} against Sackin's index. Each of these classifiers was evaluated
with 1000 two-fold cross-validations. We also performed a \gls{kPCA} projection
of the kernel matrix, and used it to visualize the separation of the different
\gls{alpha} values in the tree kernel's feature space. A schematic of this
experiment is presented in \cref{fig:kernelexpt}.

Similar experiments were performed with the values shown in
\cref{tab:kernelexpt}. The other three \gls{BA} parameters, namely \gls{N},
\gls{m}, and \gls{I}, were each varied while holding the others fixed. The
experiments for \gls{alpha}, \gls{m}, and \gls{N} were repeated with three
different values of \gls{I}. All experiments were repeated with trees having
three different numbers of tips. Kernel matrices were computed for all pairs of
the meta-parameters \gls{lambda} = \sett{0.2, 0.3, 0.4} and \gls{sigma} =
\sett{\nicefrac18, \nicefrac14, \nicefrac12, 1, 2, 4, 8}.

\begin{landscape}
\begin{table}[ht]
  \centering
  \input{\tablepath/kernel-expt}
  \caption[Variables used in tree kernel simulation experiments]
  {
    Values of parameters and other variables used in tree kernel simulation
    experiments. Each row corresponds to one of the \gls{BA} model parameters.
    One kernel matrix was created for every combination of values except the
    one indicated in the ``varied parameter'' column, which was varied when
    producing simulated trees.
  }
  \label{tab:kernelexpt}
\end{table}

\begin{table}[ht]
  \centering
  \input{\tablepath/gridsearch-expt}
  \caption[Variables used in grid search experiments]
  {
    Variables and \gls{BA} parameter values used for grid search experiments. 
    Trees were simulated under the test values, and compared to a grid of trees
    simulated under the grid values. Kernel scores were used to calculate point
    estimates and credible intervals for the test values.
  }
  \label{tab:gridexpt}
\end{table}
\end{landscape}

\begin{figure}[ht]
  \centering
  \includegraphics{kernel-expt.pdf}
  \caption[Schematic of kernel classifier experiment]{
    Schematic of investigation of \gls{BA} model parameters using the tree
    kernel.
  }
  \label{fig:kernelexpt}
\end{figure}

\subsubsection*{Grid search}

As in the previous section, we will begin by describing a single experiment,
and then list the variables for which similar experiments were performed. We
varied \gls{alpha} along a narrowly spaced grid of values: 0, 0.01, \ldots, 2.
For each value, fifteen networks were generated with \software{igraph}, and
transmission trees were simulated over each using \software{nettree}. These
trees will be referred to as ``grid trees'', and their associated values ``grid
values''. Next, one further test tree was simulated with the test value
\gls{alpha} = 0. Both the grid trees and the test tree had 500 tips, and were
simulated with the other \gls{BA} parameters set to $N$ = 5000, $m$ = 2, and
$I$ = 1000. The test tree was compared to each of the grid trees using the tree
kernel, with the meta-parameters set to $\lambda = 0.3$ and $\sigma = 4$, using
the \software{treekernel} binary. The median kernel score was calculated for
each grid value. Grid values were resampled with probability proportional to
their median kernel scores. A point estimate for the test value was obtained by
taking the highest point of an estimated kernel density of the resampled grid
values, and a 95\% credible interval was obtained using the
\software{HPDinterval} function in the \software{coda} package.

Each experiment of the type just described was repeated ten times with the same
test value. Similar experiments were performed for each of the four \gls{BA}
parameters, with several test values and trees of varying sizes. The variables
are listed in \cref{tab:gridexpt}.

\subsubsection*{Approximate Bayesian computation}

To test the full kernel-\gls{ABC} algorithm, we simulated four trees each under
a variety of parameter values, and ran the \software{netabc} program to
estimate posterior distributions for the parameters. The parameter values and
priors used are listed in \ref{tab:abcexpt}. The tree kernel meta-parameters
were set to $\lambda = 0.3$ and $\sigma = 4$. The \gls{SMC} algorithm was run
with 1000 particles, five sampled datasets per particle, and the parameter
controlling the tradeoff between speed and approximation quality
\textcite[called ``$\alpha$'' in][]{del2012adaptive} set to 0.95. The algorithm
was stopped when the acceptance rate of the \gls{MH} kernel dropped below
1.5\%, the same criterion used by \citeauthor{del2012adaptive}. Approximate
marginal posterior densities for each parameter were calculated using the
\software{density} function in \software{R} applied to the final weighted
population of particles. Credible intervals were obtained for each parameter
using the \software{HPDinterval} function in the \software{coda} package.

\begin{table}[ht]
  \centering
  \input{\tablepath/abc-expt}
  \caption[Variables used in grid search experiments]
  {
    Variables and \gls{BA} parameter values used for \gls{ABC} validation
    experiments. Trees were simulated under the test values, and
    kernel-\gls{ABC} was used to re-estimate posterior distributions for the
    \gls{BA} parameters without training.
  }
  \label{tab:abcexpt}
\end{table}

\subsubsection*{Characterization of power-law exponent in Barab\'asi-Albert networks}

Most studies of social network or transmission network
parameters~\autocite[e.g.][]{liljeros2001web, jones2003assessment,
schneeberger2004scale, brown2011transmission} report the coefficient
\gls{gamma} of the power law degree distribution. To make our results
comparable to previous work, we used simulated networks to investigate the
relationship between the \gls{BA} model parameters and \gls{gamma}. A network
was simulated for each combination of parameters listed in
\cref{tab:gammaexpt}. A power law distribution was fitted to the degree
distribution of each simulated network using the \software{fit\_power\_law}
function in \software{igraph} with the `R.mle' implementation. We fitted a
\gls{GLM} with Gamma-distributed errors and a log link function to the observed
distribution of \gls{gamma} values, with \gls{alpha}, \gls{m}, \gls{N}, and all
possible interaction terms as predictors.

\begin{table}
  \centering
  \input{\tablepath/gamma-expt}
  \caption[\gls{BA} parameters used as input \gls{GLM} predicting $\gamma$]
  {
    \gls{BA} model parameters used as input to \gls{GLM} predicting power law
    exponent $\gamma$. One network was simulated with each combination of
    parameters, and $\gamma$ was calculated for each network. A \gls{GLM} with
    Gamma-distributed errors and a log link function was fit to the $\gamma$
    values with all parameters and interaction terms as predictors.
  }
  \label{tab:gammaexpt}
\end{table}

@@

Because the \gls{BA} model assumes a single connected contact network, it is
most appropriate to apply to groups of individuals who are epidemiologically
related. Therefore, we searched for published \gls{HIV} datasets which
originated from existing clusters, either phylogenetically or geographically
defined. In addition, we analysed an in-house dataset sampled from
\gls{HIV}-positive individuals in British Columbia, Canada (the ``BC data'').
The datasets are summarized in \cref{tab:realdata}.

We downloaded all sequences associated with each published study from GenBank.
For the \textcite{novitsky2014impact} data, each sequence was aligned pairwise
to the HXB2 reference sequence (GenBank accession number HIVHXB2CG) and the
hypervariable regions were clipped out with \software{BioPython} version
1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
\software{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments
were manually inspected with \software{Seaview} version 4.4.2
\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
alignments by approximate maximum likelihood using \software{FastTree2} version
2.1.7 with the \gls{GTR} model. Transmission trees were estimated by rooting
and time-scaling the phylogenies by root-to-tip regression, using a modified
version of Path-O-Gen (distributed as part of
BEAST~\autocite{drummond2007beast}) as described
previously~\autocite{poon2015phylodynamic}.

Three of the datasets \autocite[][and the BC data]{li2015hiv,novitsky2014impact}
were initially much larger than the others, containing 1265, 1299, and 7923
sequences respectively. To ensure that the analyses were comparable, we reduced
these to a number of sequences similar to the smaller datasets. For the
\citeauthor{li2015hiv} and BC datasets, we detected clusters of size 280 and
399 respectively using a patristic distance cutoff of 0.02 as described
previously~\autocite{poon2014impact}. Only sequences within these clusters were
carried forward. For the \textcite{novitsky2014impact} data, no large clusters
were detected using the same cutoff, so we analysed a subtree of size 180
chosen arbitrarily.

\begin{table}[ht]
  \centering
  \input{\tablepath/realdata-bc}
  \caption{Characteristics of published HIV datasets analyzed with kernel-ABC.}
  \label{tab:realdata}
\end{table}

### 2016-03-31 doc/thesis/results.Rnw
Finished results! Yes!

% overview of experiment
Our first simulation-based experiment was designed to determine which network
model parameters were potentially estimable by kernel-ABC, and to find the best
set of kernel meta-parameters to carry through to future experiments. For each
of the four \gls{BA} model parameters, we simulated trees under three distinct
values of that parameter while holding the others fixed, then cross-validated a
kernel-\gls{SVM} classifier for that parameter.

<<treestats, include=FALSE>>=
    params <- c("alpha", "I", "m", "N")
    ptn <- paste0("../../simulations/kernel-", params, "/statistics/*")
    files <- lapply(ptn, Sys.glob)
    d <- lapply(files, collect.data)
    d <- mapply(cbind, d, parameter=as.list(params), SIMPLIFY=FALSE)
    lapply(d, setDT)
    d[[1]] <- subset(d[[1]], m == 2)
    setnames(d[[1]], "nsimnode", "I")
@

% trees look different
Trees simulated under different values of \gls{alpha} are visibly quite
distinctive (\cref{fig:patrees}). In particular, higher values of \gls{alpha}
produce networks with a small number of highly connected nodes which, once
infected, are likely to transmit to many other nodes (\cref{fig:alphabds}).
This results in a more unbalanced, ladder-like structure in the phylogeny,
compared to networks with lower \gls{alpha} values. None of the other three
parameters produced trees which were as easily distinguished from each other
(\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}). However, higher $I$
values produce trees which appear to have longer terminal branches,

@@

The fact that Sackin's index is informative of \gls{alpha} was
unsurprising, given the clear differences in tree balance observed under
different \gls{alpha} values.

\begin{figure}[ht]
  \centering
  \includegraphics{kernel-alpha-tree.pdf}
  \caption[Visibly distinctive trees simulated under three values of \gls{alpha}]{
    Epidemics simulated on \gls{BA} networks of 5000 nodes, with \gls{alpha}
    equal to 0.5, 1.0, or 1.5, until 1000 individuals were infected.
    Transmission trees were created by sampling 500 infected nodes. Higher
    \gls{alpha} values produced networks with a small number of
    highly-connected nodes, resulting in highly unbalanced, ladder-like trees.
  }
  \label{fig:patrees}
\end{figure}

>>>>

<<treestats, include=FALSE>>=
    params <- c("alpha", "I", "m", "N")

    get.treestats <- function (param) {
        stats <- paste0("../../simulations/kernel-", param, "/statistics/*")
        d <- collect.data(stats)
        trees <- paste0("../../simulations/kernel-", param, "/tree/*")
        m <- collect.metadata(trees)
        rownames(m) <- sub("../../simulations/", "", rownames(m))
        d <- merge(d, m, by=0, suffixes=c("", ".1"))
        if (param == "alpha") {
            d <- subset(d, m == 2)
            colnames(d)[colnames(d) == "nsimnode"] <- "I"
        }
        setDT(d)
    }

    alpha.stats <- get.treestats("alpha")
    I.stats <- get.treestats("I")
    m.stats <- get.treestats("m")
    N.stats <- get.treestats("N")

    alpha.sackin.cor <- alpha.stats[,cor.test(alpha, sackin, method="spearman")]
    stopifnot(alpha.sackin.cor$p.value < 1e-5)

    N.ratio.cor <- N.stats[,cor.test(int.tip.ratio, N)]
    I.ratio.cor <- I.stats[,cor.test(int.tip.ratio, I)]
    stopifnot(N.ratio.cor$p.value < 1e-5)
    stopifnot(I.ratio.cor$p.value < 1e-5)
@

Trees simulated under different values of \gls{alpha} are visibly quite
distinct (\cref{fig:alphatrees}). In particular, higher values of \gls{alpha}
produce networks with a small number of highly connected nodes which, once
infected, are likely to transmit to many other nodes (\cref{fig:alphabds}).
This results in a more unbalanced, ladder-like structure in the phylogeny,
compared to networks with lower \gls{alpha} values. Sackin's index was
significantly correlated with \gls{alpha} 
    (Spearman's rho \Sexpr{round(alpha.sackin.cor$estimate, 2)},
     $p < 10^{-5}$).
None of the other three parameters produced trees which were as easily
distinguished from each other
(\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}). However, the ratio of
internal to terminal branch lengths was positively correlated with $N$ and
negatively correlated with $I$ (Spearman's rho
    \Sexpr{round(N.ratio.cor$estimate, 2)} and
    \Sexpr{round(I.ratio.cor$estimate, 2)},
    both $p < 1e-5$).

@@

\begin{figure}[ht]
  \centering
  \includegraphics{kernel-alpha-tree.pdf}
  \caption[Visibly distinctive trees simulated under three values of \gls{alpha}]{
    Epidemics simulated on \gls{BA} networks of 5000 nodes, with \gls{alpha}
    equal to 0.5, 1.0, or 1.5, until 1000 individuals were infected.
    Transmission trees were created by sampling 500 infected nodes. Higher
    \gls{alpha} values produced networks with a small number of
    highly-connected nodes, resulting in highly unbalanced, ladder-like trees.
  }
  \label{fig:alphatrees}
\end{figure}
